MSCA 31006: Final Project - AQI Forecasting

### 2023-08-16 MSCA_31006_Time_Series_Final_Combined.Rmd
Add files via upload

---
title: 'MSCA 31006: Final Project - AQI Forecasting'
author: "Aashish Singh, Jacob Brewer, Nikitha Gopal, Sydney Peters"
date: "8/10/2023"
output: html_document
---
```{r}
# List all objects in the environment
objects_to_remove <- ls()

# Remove all objects
rm(list = objects_to_remove)
```

@@

```{r, tidy=TRUE}
path <- "/Users/nikithagopal/Desktop/MSCA Data/Time Series Final/"
pm2_5_data <- read.csv(paste0(path, "la_pm2_5_pollutant_data.csv"), 
                     na.strings=c("", "NA"))
ozone_data <- read.csv(paste0(path, "la_ozone_pollutant_data.csv"), 
                     na.strings=c("", "NA"))

#path <- "/Users/sydneypeters/Desktop/"
#pm2_5_data <- read.csv(paste0(path, "pm2.5_LA.csv"), 
#                     na.strings=c("", "NA"))
#ozone_data <- read.csv(paste0(path, "ozone_LA.csv"), 
#                     na.strings=c("", "NA"))

@@

#### Step 6a: Split data into train and test - Weekly

@@

#### Step 7a: Decompose the time series plot - Weekly

@@

#### Step 8a: Benchmark using snaive model - Weekly

```{r, tidy=TRUE}
# Forecast the seasonal naïve for (2021-01 to 2023-07)
forecast_snaive_weekly <- snaive(train_weekly, h=110)
# Plot the forecasts for snaive model
plot(forecast_snaive_weekly, main = "PM 2.5 AQI Forecast - SNaive (Weekly)",
         xlab = "Week", ylab = "PM 2.5 AQI")
lines(test_weekly)

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Mean Absolute Percentage Error (MAPE)
mape_snaive_weekly <- mean(abs((test_weekly - forecast_snaive_weekly$mean)/test_weekly))
mape_snaive_weekly

@@

# Mean Squared Error (MSE)
mse_snaive_weekly <- mean((test_weekly - forecast_snaive_weekly$mean)^2)
mse_snaive_weekly

@@

#### Step 5b: Extract values for time series and plot the series - Monthly

```{r, tidy=TRUE}
# Create monthly data
pm2_5_monthly <- apply.monthly(pm2_5_xts, mean)

pm2_5_monthly_ts <- ts(pm2_5_monthly["19990131/20230811"], start = c(1999,1), frequency = 12)
plot(pm2_5_monthly_ts)

kpss.test(pm2_5_monthly_ts)

adf.test(pm2_5_monthly_ts)

acf(pm2_5_monthly_ts, main = "ACF of Monthly Time Series")
pacf(pm2_5_monthly_ts, main = "PACF of Monthly Time Series")

#### Step 6b: Split data into train and test - Monthly

```{r, tidy=TRUE}
# Split the data into a training and test dataset
train_monthly <- window(pm2_5_monthly_ts, start = c(1999,1), end=c(2020,12))
test_monthly <- window(pm2_5_monthly_ts, start = c(2021,1))
```

#### Step 7b: Decompose the time series plot - Monthly

# Looking at spectral analysis
periodogram(train_monthly, log = "no", plot=TRUE, 
            ylab = "Periodogram",
            xlab = "Frequency",
            lwd=2, xlim = c(0, 0.2))

# There are two trends: 
#  1) A typical yearly (12 months) seasonal trend
#  2) A trend that is repeating every 8-9 years (100 months)
#  3) A typical half yearly (6 months) seasonal trend

# Decompose the series
plot(decompose(train_monthly, type="multiplicative"))

# Understand the seasonality and remove it to see the trend
train_monthly_diff <- diff(train_monthly, lag = 12)
autoplot(train_monthly_diff, ylab="Train Seasonal Differencing (Monthly)")

# Let's look if the series is stationary?
kpss.test(train_monthly_diff)
# The reported p-value is 0.1, which is > 0.05, and would suggest that we fail 
# to reject the null hypothesis of level stationarity (conclusion: stationary)

adf.test(train_monthly_diff)
# The reported p-value of 0.01 (<0.05) so we do reject the null hypothesis that 
# the time series is non-stationary (conclusion: stationary)

acf(train_monthly_diff, main = "ACF of Seasonal Differencing Time Series")
pacf(train_monthly_diff, main = "PACF of Seasonal Differencing Time Series")

#### Step 8b: Benchmark using snaive model - Monthly

# Forecast the seasonal naïve for (2021-01 to 2023-07)
forecast_snaive_monthly <- snaive(train_monthly, h=31)
# Plot the forecasts for snaive model
plot(forecast_snaive_monthly, main = "PM 2.5 AQI Forecast - SNaive",
         xlab = "Week", ylab = "PM 2.5 AQI")
lines(test_monthly)

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Mean Absolute Percentage Error (MAPE)
mape_snaive <- mean(abs((test_monthly - forecast_snaive_monthly$mean)/test_monthly))
mape_snaive

# Mean Squared Error (MSE)
mse_snaive <- mean((test_monthly - forecast_snaive_monthly$mean)^2)
mse_snaive

# Evaluate the residuals
checkresiduals(forecast_snaive_monthly)

#### Step 9a: TBATS model - Weekly

```{r, tidy=TRUE}
vec <- as.vector(train_weekly)

univariate_ts <- ts(vec, start=1999, frequency=52.18)

stl_decomposition <- stl(univariate_ts, s.window="periodic")
plot(stl_decomposition)

h <- length(test_weekly)

model_tbats <- tbats(train_weekly)
forecast_tbats <- forecast(model_tbats, h=h)
plot(forecast_tbats)

checkresiduals(model_tbats)
#Ljung-Box test .05179

plot(pm2_5_weekly_ts, xlim=c(1999, 2023), ylim=c(min(pm2_5_weekly_ts), max(pm2_5_weekly_ts)), main="Train, Test, and Forecasted Data - Weekly", xlab="Year", ylab="PM2.5 AQI Value")
lines(test_weekly, col="red") # Test data in red
lines(forecast_tbats$mean, col="blue", type="o") # Forecasted data in blue
legend("topleft", legend=c("Train", "Test", "Forecast"), fill=c("black", "red", "blue"))

mae <- mean(abs(test_weekly - forecast_tbats$mean))
print(cat("Mean Absolute Error (MAE):", round(mae, 2), "\n"))
#Mean Absolute Error (MAE): 9.31

comparison_df <- data.frame(
  Date = time(test_weekly),
  Actual = as.vector(test_weekly),
  Forecasted = as.vector(forecast_tbats$mean)
)
print(comparison_df)

#### Step 9b: TBATS model - Monthly

vec <- as.vector(train_monthly)

univariate_ts <- ts(vec, start=1999, frequency=52.18)

stl_decomposition <- stl(univariate_ts, s.window="periodic")
plot(stl_decomposition)

h <- length(test_monthly)

model_tbats <- tbats(train_monthly)
forecast_tbats <- forecast(model_tbats, h=h)
plot(forecast_tbats)

checkresiduals(model_tbats)
#Ljung-Box test .07956

plot(pm2_5_monthly_ts, xlim=c(1999, 2023), ylim=c(min(pm2_5_monthly_ts), max(pm2_5_monthly_ts)), main="Train, Test, and Forecasted Data - Monthly", xlab="Year", ylab="PM2.5 AQI Value")
lines(test_monthly, col="red") # Test data in red
lines(forecast_tbats$mean, col="blue", type="o") # Forecasted data in blue
legend("topleft", legend=c("Train", "Test", "Forecast"), fill=c("black", "red", "blue"))

mae <- mean(abs(test_monthly - forecast_tbats$mean))
print(cat("Mean Absolute Error (MAE):", round(mae, 2), "\n"))
#Mean Absolute Error (MAE): 6.61

comparison_df <- data.frame(
  Date = time(test_monthly),
  Actual = as.vector(test_monthly),
  Forecasted = as.vector(forecast_tbats$mean)
)
print(comparison_df)

#### Step 10a: ARIMA model using auto.arima() - Weekly
```{r}
eacf <-
function (z,ar.max=7,ma.max=13) 
{
#
#  PROGRAMMED BY K.S. CHAN, DEPARTMENT OF STATISTICS AND ACTUARIAL SCIENCE,
#  UNIVERSITY OF IOWA.
#
#  DATE: 4/2001
#  Compute the extended sample acf (ESACF) for the time series stored in z.
#  The matrix of ESACF with the AR order up to ar.max and the MA order
#  up to ma.max is stored in the matrix EACFM.
#  The default values for NAR and NMA are 7 and 13 respectively.
#  Side effect of the eacf function:
#  The function prints a coded ESACF table with
#  significant values denoted by * and nosignificant values by 0, significance
#  level being 5%.
#
#  Output:
#   eacf=matrix of esacf
#   symbol=matrix of coded esacf
#

@@

```{r}

kpss_test_result <- kpss.test(train_weekly_diff)
adf_test_result <- adf.test(train_weekly_diff)

cat("KPSS: ", round(kpss_test_result$p.value, 2), "\n")
cat("ADF: ", round(adf_test_result$p.value, 2), "\n")

# Fit ARIMA model using auto.arima()
arima_model_weekly <- auto.arima(train_weekly, seasonal = TRUE, stepwise = FALSE,
                                 trace = FALSE,
                                 start.p = 1, max.p = 4,
                                 start.q = 2, max.q = 5)

# auto.arima(pm2_5_weekly_diff, lambda = "auto", seasonal=TRUE, )

# Forecast using ARIMA model
h <- length(test_weekly)
forecast_arima_weekly <- forecast(arima_model_weekly, h = h)

plot(forecast_arima_weekly, main = "PM 2.5 AQI Forecast - sARIMA",
         xlab = "Week", ylab = "PM 2.5 AQI")
lines(test_weekly)

@@

```{r}

# Check the residuals of the ARIMA model
checkresiduals(arima_model_weekly)

# Calculate MAE
mae <- mean(abs(forecast_arima_weekly$mean - test_weekly))
# Calculate MAPE
mape <- mean(abs((forecast_arima_weekly$mean - test_weekly) / test_weekly)) * 100

# Get AICc and BIC values from the ARIMA model
aic <- arima_model_weekly$aic
bic <- arima_model_weekly$bic

# Corrected AICc calculation
n <- length(train_weekly)
k <- length(arima_model_weekly$coef)
aic_corrected <- aic + 2 * (k + 1) * (k + 2) / (n - k - 2)

# Print the results
cat("MAE: ", round(mae, 2), "\n")
cat("MAPE: ", round(mape, 2), "%\n")
cat("AICc: ", round(aic_corrected, 2), "\n")
cat("BIC: ", round(bic, 2), "\n")

# Create a time index for the forecasted values
forecast_time_index <- seq(from = end(test_weekly) + 1 / 52, 
                           length.out = length(forecasted_values), by = 1 / 52)

# Plot historical data, actual test, and forecast
plot(train_weekly, xlim=c(1999, 2023), ylim=c(min(train_weekly), max(train_weekly)), main="Train, Test, and Forecasted Data - Weekly", xlab="Year", ylab="PM2.5 AQI Value")
lines(test_weekly, col="red") # Test data in red
lines(forecast_arima_weekly$mean, col="blue", type="o") # Forecasted data in blue
legend("topleft", legend=c("Train", "Test", "Forecast"), fill=c("black", "red", "blue"))

comparison_df <- data.frame(
  Date = time(test_weekly),
  Actual = as.vector(test_weekly),
  Forecasted = as.vector(forecast_arima_weekly$mean)
)
print(comparison_df)
```
```{r}
summary(arima_model_weekly)
```

```{r}
# Assuming your residuals are stored in a variable called "residuals"
shapiro_test_result <- shapiro.test(arima_model_weekly$residuals)

# Print the test result
print(shapiro_test_result)

# Check the p-value
if (shapiro_test_result$p.value < 0.05) {
  cat("The residuals are not normally distributed (p-value:", shapiro_test_result$p.value, ")\n")
} else {
  cat("The residuals appear to be normally distributed (p-value:", shapiro_test_result$p.value, ")\n")
}
```

>>>>

---
title: 'MSCA 31006: Final Project - AQI Forecasting'
author: "Aashish Singh, Jacob Brewer, Nikitha Gopal, Sydney Peters"
date: "8/10/2023"
output: html_document
---

@@

```{r, tidy=TRUE}
path <- "/Users/aashishsingh/Desktop/Time Series - Final Project/"
pm2_5_data <- read.csv(paste0(path, "la_pm2_5_pollutant_data.csv"), 
                     na.strings=c("", "NA"))
ozone_data <- read.csv(paste0(path, "la_ozone_pollutant_data.csv"), 
                     na.strings=c("", "NA"))

@@

#### Step 6: Split data into train and test - Weekly

@@

#### Step 7: Decompose the time series plot - Weekly

@@

#### Step 8: Benchmark using snaive model - Weekly

```{r, tidy=TRUE}
# Forecast the seasonal naïve for (2021-01 to 2023-07)
h <- length(test_weekly)
forecast_snaive_weekly <- snaive(train_weekly, lambda="auto", h)
summary(forecast_snaive_weekly)
# Plot the forecasts for snaive model
plot(forecast_snaive_weekly, main = "PM 2.5 AQI Forecast - SNaive (Weekly)",
         xlab = "Week", ylab = "PM 2.5 AQI")
lines(test_weekly)

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Symmetric Mean Absolute Percentage Error (sMAPE)
smape_snaive_weekly <- smape(test_weekly, forecast_snaive_weekly$mean)
smape_snaive_weekly

@@

# Root Mean Squared Error (RMSE)
rmse_snaive_weekly <- rmse(test_weekly, forecast_snaive_weekly$mean)
rmse_snaive_weekly

@@

#### Step 9: Forecast using SARIMA - Weekly

```{r}
eacf <-
function (z,ar.max=7,ma.max=13) 
{
#
#  PROGRAMMED BY K.S. CHAN, DEPARTMENT OF STATISTICS AND ACTUARIAL SCIENCE,
#  UNIVERSITY OF IOWA.
#
#  DATE: 4/2001
#  Compute the extended sample acf (ESACF) for the time series stored in z.
#  The matrix of ESACF with the AR order up to ar.max and the MA order
#  up to ma.max is stored in the matrix EACFM.
#  The default values for NAR and NMA are 7 and 13 respectively.
#  Side effect of the eacf function:
#  The function prints a coded ESACF table with
#  significant values denoted by * and nosignificant values by 0, significance
#  level being 5%.
#
#  Output:
#   eacf=matrix of esacf
#   symbol=matrix of coded esacf
#

@@

```{r, tidy=TRUE}
# Forecast the TBATS for (2021-01 to 2023-07)
h <- length(test_weekly)

model_sarima_weekly <- auto.arima(train_weekly, 
                                  seasonal = TRUE, 
                                  stepwise = FALSE, #stepwise = TRUE,
                                  trace = FALSE,
                                  #m=52,
                                  #d=None,
                                  #D=1, # order of seasonal differencing
                                  lambda="auto",
                                  start.p = 1, max.p = 3,
                                  start.q = 2, max.q = 4)

forecast_sarima_weekly <- forecast(model_sarima_weekly, h)
plot(forecast_sarima_weekly)

# Plot the forecasts for TBATS model
plot(pm2_5_weekly_ts, 
     xlim=c(1999, 2023), 
     ylim=c(min(pm2_5_weekly_ts), 
            max(pm2_5_weekly_ts)), 
     main="Train, Test, and Forecasted Data - Weekly", 
     xlab="Year", 
     ylab="PM2.5 AQI Value")
lines(test_weekly, col="red") # Test data in red
lines(forecast_sarima_weekly$mean, col="blue", type="o") # Forecasted data in blue
legend("topleft", 
       legend=c("Train", "Test", "Forecast"), 
       fill=c("black", "red", "blue"))

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Symmetric Mean Absolute Percentage Error (sMAPE)
smape_sarima_weekly <- smape(test_weekly, forecast_sarima_weekly$mean)
smape_sarima_weekly

# Mean Absolute Error (MAE)
mae_sarima_weekly <- mean(abs((test_weekly - forecast_sarima_weekly$mean)))
mae_sarima_weekly

# Root Mean Squared Error (RMSE)
rmse_sarima_weekly <- rmse(test_weekly, forecast_sarima_weekly$mean)
rmse_sarima_weekly

# Evaluate the residuals
checkresiduals(forecast_sarima_weekly)

comparison_df <- data.frame(
  Date = time(test_weekly),
  Actual = as.vector(test_weekly),
  Forecasted = as.vector(forecast_sarima_weekly$mean)
)
print(comparison_df)

@@

#### Step 10: Forecast using TBATS - Weekly

```{r, tidy=TRUE}
# Forecast the TBATS for (2021-01 to 2023-07)
h <- length(test_weekly)

vec <- as.vector(train_weekly)
univariate_ts <- ts(vec, start=1999, frequency=52.18)
plot(stl(univariate_ts, s.window="periodic"))

model_tbats_weekly <- tbats(train_weekly,
                            use.box.cox = TRUE)
forecast_tbats_weekly <- forecast(model_tbats_weekly, h)
plot(forecast_tbats_weekly)

# Plot the forecasts for TBATS model
plot(pm2_5_weekly_ts, 
     xlim=c(1999, 2023), 
     ylim=c(min(pm2_5_weekly_ts), 
            max(pm2_5_weekly_ts)), 
     main="Train, Test, and Forecasted Data - Weekly", 
     xlab="Year", 
     ylab="PM2.5 AQI Value")
lines(test_weekly, col="red") # Test data in red
lines(forecast_tbats_weekly$mean, col="blue", type="o") # Forecasted data in blue
legend("topleft", 
       legend=c("Train", "Test", "Forecast"), 
       fill=c("black", "red", "blue"))

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Symmetric Mean Absolute Percentage Error (sMAPE)
smape_tbats_weekly <- smape(test_weekly, forecast_tbats_weekly$mean)
smape_tbats_weekly

# Mean Absolute Error (MAE)
mae_tbats_weekly <- mean(abs((test_weekly - forecast_tbats_weekly$mean)))
mae_tbats_weekly

# Root Mean Squared Error (RMSE)
rmse_tbats_weekly <- rmse(test_weekly, forecast_tbats_weekly$mean)
rmse_tbats_weekly

# Evaluate the residuals
checkresiduals(forecast_tbats_weekly)

comparison_df <- data.frame(
  Date = time(test_weekly),
  Actual = as.vector(test_weekly),
  Forecasted = as.vector(forecast_tbats_weekly$mean)
)
print(comparison_df)

#### Step 11: Forecast using Neural Network - Weekly

```{r, tidy=TRUE}
h <- length(test_weekly)
model_nn_weekly <- nnetar(train_weekly,
                           size=10,
                           decay=0.5,
                           repeats=1000,
                           lambda="auto")
summary(model_nn_weekly)
forecast_nn_weekly <- forecast(model_nn_weekly, PI=FALSE, h)
plot(forecast_nn_weekly)

# Plot the forecasts for NN model
plot(pm2_5_weekly_ts, 
     xlim=c(1999, 2023), 
     ylim=c(min(pm2_5_weekly_ts), 
            max(pm2_5_weekly_ts)), 
     main="Train, Test, and Forecasted Data - Weekly", 
     xlab="Year", 
     ylab="PM2.5 AQI Value")
lines(test_weekly, col="red") # Test data in red
lines(forecast_nn_weekly$mean, col="blue", type="o") # Forecasted data in blue
legend("topleft", 
       legend=c("Train", "Test", "Forecast"), 
       fill=c("black", "red", "blue"))

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Symmetric Mean Absolute Percentage Error (sMAPE)
smape_nn_weekly <- smape(test_weekly, forecast_nn_weekly$mean)
smape_nn_weekly

# Mean Absolute Error (MAE)
mae_nn_weekly <- mean(abs((test_weekly - forecast_nn_weekly$mean)))
mae_nn_weekly

# Root Mean Squared Error (RMSE)
rmse_nn_weekly <- rmse(test_weekly, forecast_nn_weekly$mean)
rmse_nn_weekly

# Evaluate the residuals
checkresiduals(forecast_nn_weekly)

comparison_df <- data.frame(
  Date = time(test_weekly),
  Actual = as.vector(test_weekly),
  Forecasted = as.vector(forecast_nn_weekly$mean)
)
print(comparison_df)
